---
title: "Ende-TAM"
author: "Kate Lindsey"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data mining

The raw data comes from a .txt file called phrases.txt, which was provided by Jack Rabinovitch. The data was scraped from an April 2023 version of the Ende FLEx corpus.

```{r}
library(readr)
phrases <- read_delim("C:/Users/profk/Downloads/phrases.txt", 
     delim = "\t", escape_double = FALSE, 
     trim_ws = TRUE)
```

This dataframe has 37,772 phrases.

To focus on phrases with verbs, I reduced the data to only the rows that have "verb" in the words_pos column. This should capture all words marked as synthetic/analytic verbs, auxiliary verbs, and copular verbs.

```{r}
df_TAM <- subset(phrases, grepl("verb", words_pos))
```

The new dataframe has 20,355 phrases.

To tidy up the dataframe, I will create a new column for the phrase (phrase_txt), which is words_txt without the commas. I will rename the gls column to phrase_tr, the words_txt to word_txt, the words_gl to word_tr, the words_pos to word_pos and I will remove all columns except for word_tr, word_txt, phrase_tr, and phrase_txt.

```{r}
library(dplyr)

tidy_TAM <- df_TAM %>% mutate(phrase_txt = gsub(",", " ", words_txt))
tidy_TAM$phrase_txt <- gsub("  ", " ", tidy_TAM$phrase_txt)
tidy_TAM$phrase_txt <- gsub("  ", " ", tidy_TAM$phrase_txt)
tidy_TAM$phrase_txt <- gsub("  ", " ", tidy_TAM$phrase_txt)
tidy_TAM <- tidy_TAM %>% rename(phrase_tr = gls, word_txt = words_txt, word_tr = words_gls, word_pos = words_pos)
tidy_TAM <- select(tidy_TAM, word_txt, word_tr, word_pos, phrase_txt, phrase_tr)
```
Now, we will reorganize the dataframe so that each row represents one word.

```{r}
library(tidyr)
library(stringr)

tidy_TAM$word_tr <- gsub("correct, true, right", "correct", tidy_TAM$word_tr) 
tidy_TAM$word_tr <- gsub("Idi borrowing for ll, to go", "to go", tidy_TAM$word_tr)
#this translation was giving us trouble

tidy_TAM <- tidy_TAM %>% 
  mutate(
    word_txt_commas = stringr::str_count(word_txt, ", "),
    word_tr_commas = stringr::str_count(word_tr, ", ")
  ) %>% 
  mutate(
    word_tr = ifelse(word_tr_commas < word_txt_commas, paste0(word_tr, ", "), word_tr)
  )

tidy_TAM <- tidy_TAM %>% 
  mutate(
    word_tr_commas = stringr::str_count(word_tr, ", "),
    word_pos_commas = stringr::str_count(word_pos, ", ")
  ) %>% 
  mutate(
    word_pos = ifelse(word_pos_commas < word_tr_commas, paste0(word_pos, ", "), word_pos)
  )

tidy_TAM <- tidy_TAM %>% separate_rows(word_txt, word_tr, word_pos, sep = ", ")
tidy_TAM <- tidy_TAM %>% mutate_all(str_trim)
```

Now, we will clean up the rows once more, only retaining the words with "verb" in the word_pos column, removing capitalization and commas, and removing the count commas columns.

```{r}
tidy_TAM <- subset(tidy_TAM, grepl("verb", word_pos))
tidy_TAM <- tidy_TAM[!grepl("adverb", tidy_TAM$word_pos), ]
tidy_TAM$word_txt <- tolower(tidy_TAM$word_txt)
tidy_TAM$word_txt <- gsub(",", "", tidy_TAM$word_txt)
tidy_TAM$word_tr <- gsub(",", "", tidy_TAM$word_tr)
tidy_TAM$word_pos <- gsub(",", "", tidy_TAM$word_pos)
tidy_TAM <- tidy_TAM %>% 
  select(-word_txt_commas, -word_tr_commas, -word_pos_commas)
```

Now, we will search the word_tr column for any tense, aspect, or modality information and puts it into a column verb_TAM.

```{r}
tidy_TAM <- tidy_TAM %>%
  mutate(verb_TAM = case_when(
    str_detect(word_tr, "\\.prs\\.") ~ "prs",
    str_detect(word_tr, "\\.rem\\.") ~ "rem",
    str_detect(word_tr, "\\.rec\\.") ~ "rec",
    str_detect(word_tr, "\\.fut\\.") ~ "fut",
    str_detect(word_tr, "\\.irr\\.") ~ "irr",
    str_detect(word_tr, "\\.cond\\.") ~ "cond",
    TRUE ~ NA_character_
  ))
```

We will now reduce the dataframe to those verbs that have TAM information.

```{r}
tidy_TAM <- tidy_TAM[complete.cases(tidy_TAM$verb_TAM), ]
```

Now, we will add two columns, one for previous verb TAM, one for following verb TAM information, and one that concatenates the three together.

```{r}
tidy_TAM$prev_verb_TAM <- lag(tidy_TAM$verb_TAM)
tidy_TAM$next_verb_TAM <- lead(tidy_TAM$verb_TAM)
tidy_TAM <- tidy_TAM %>%
  mutate(TAM_pattern = paste(prev_verb_TAM, verb_TAM, next_verb_TAM, sep = "."))
```


